The art of modelling the 3D structure of the genome\\ Insights in {\em P. falciparum} genome organization

### 2017-07-26 doc/manuscript/paper.tex
more figures and schemas - rewording the introduction

% Title of paper
\title{Analysing the three-dimensional structure of {\em P. falciparum}}
\author{Nelle Varoquaux}

@@

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependant}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}.

Within the past few years, two studies of the three-dimensional structure of
{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
several strains of {\em P. falciparum} to understand the link between 3D
structure and the complex regulation of \textit{var} genes, a family of genes
involved the invasion of red blood cell and also responsible for the
parasite's great capacity to evade our immune system.
\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
parasite during three key stages of the erythrocytic cycle.
\citet{ay:multiple} reviews the key biological findings of those studies.

An important part of analysing the three-dimensional structure of the
genome of {\em P. falciparum} (as well as many other organisms) relied on the
inferring accurate 3D models of the structure. These models were then used as
a preliminary step to various analysis, such as identifying colocolized
elements, distinguishing between open and closed chromatin and last but not
least, intergrating with other sources of data, such as gene expression or
chromatin modification.

@@

@@

The development of genowe-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA.

Over the past 10 years, the challenge of building 3D models from contact maps
have been tackled by a plethora of methods. These model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

@@

The objective function $\mathcal{O}$, sometimes called the ``scoring
function''

@@

The first step is thus to find an adequate count-to-distance mapping to
convert contact counts into pairwise wish-distances $\delta_{ij}$. 
\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
cerevisiae}'s chromosome III as a Worm-like chain polymer,
\citet{duan:three-dimensional} uses a linear mapping between contact counts
and wish distances, \citet{ay:three-dimensional} relies on the biophysical
properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
count-to-distance mapping by fitting the relationship using known pairwise
distances obtained through high-resolution FISH measures.

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio.

@@

\citet{paulsen:manifold} proposes a very different method, and formulate a
\textit{manifold based optimization}, where a low rank PSD matrix (and thus a
distance matrix) is optimized to be as close as possible to the sparse contact
count matrix. Applying classical MDS on this low rank PSD matrix then yields a
3D model of the genome.

@@

In the previous section, we discussed how from a contact count map one could
infer a {\em consensus} model or a population of models that best represented
the contact map. Another set of methods approach the problem of finding 3D
models of genome architecture very differently. Instead of using data-driven
approaches, these methods model chromosomes as series of beads or flexible
fibers, and attempt to find the smallest set of constraints such that a
generated contact maps from these models match as closely as possible the
observed contact maps. As these models refine the constraints, more insight on
the hallmarks of the genome architecture are discovered.

>>>>

% Title of paper
\title{The art of modelling the 3D structure of the genome\\
  Insights in {\em P. falciparum} genome organization}
\author{Nelle Varoquaux}

@@

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}. Thus,
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependent}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
years, two studies of the three-dimensional structure of {\em P. falciparum}
have been released.  \citet{lemieux:genome-wide} probed several strains of
{\em P. falciparum} to understand the link between 3D structure and the
complex regulation of \textit{var} genes, a family of genes involved the
invasion of red blood cell and also responsible for the parasite's great
capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
3D structure of the genome of the parasite during three key stages of the
erythrocytic cycle. The relatively small size of the genome yet relatively
complicated genome architecture, the complex life cycle and the strong link
between chromatine folding, gene regulation and epigenetic places {\em P.
falciparum} as an interesting case study for developing new methods for better
understanding the properties of genome folding and its link to gene
regulation. An important part of analysing the three-dimensional structure of
the genome of {\em P. falciparum} (as well as many other organisms) relied on
the inferring accurate 3D models of the structure. These models were then used
as a preliminary step to various analysis, such as identifying colocolized
elements, distinguishing between open and closed chromatin and last but not
least, intergrating with other sources of data, such as gene expression or
chromatin modification.

\fixme{Add why we are doing a review}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
\caption{The art of understanding 3D genome structure}{
\fixme{write caption}}
\label{Fig1}
\end{figure}

@@

The development of genowe-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA. Over the past 10 years, the
challenge of building 3D models from contact maps have been tackled by a
plethora of methods, some developped for a particular organisms, others with
the hope to be generalizable to any data sets. Methods model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

\fixme{contrast model-driven studies with data-driven studies}

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/counts_maps.png}
\caption{Contact maps of {\em P. falciparum}'s chr7}{
\fixme{write caption}
\textbf{A.} \citep{lemieux:genome-wide}}
\label{Fig1}
\end{figure}

@@

The objective function $\mathcal{O}$, sometimes called the ``scoring
function'', can be derived from known embedding algorithms (such as
multidimensional-scaling methods), from statistical modeling of contact
counts, or simply constructed in an \textit{ad hoc} manner to fulfill a set of
wished properties.

@@

The first step is thus to find an adequate count-to-distance mapping to
convert contact counts into pairwise wish-distances $\delta_{ij}$. 
\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
cerevisiae}'s chromosome III as a Worm-like chain polymer,
\citet{duan:three-dimensional} uses a linear mapping between contact counts
and wish distances, \citet{ay:three-dimensional} relies on the biophysical
properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
count-to-distance mapping by fitting the relationship using known pairwise
distances obtained through high-resolution FISH measures. In brief, there are
as many ways to derive the count-to-distance mapping as there are of studies.

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio, through a direct modeling of the
noise through the statistical model.

@@

\citet{paulsen:manifold} proposes a very different method, and formulate a
\textit{manifold based optimization}, where a low rank PSD matrix (and thus a
distance matrix) is optimized to be as close as possible to the sparse contact
count matrix. Applying classical MDS on this low rank PSD matrix then yields a
\textit{consensus} 3D model of the genome.

@@

``data-driven'' methods, as presented in the last section, use the
experimental contact maps to infer models as consistent as possible with the
data. Another approach consists in modeling chromosomes as polymers under a
small number constraints such that a generated contact maps from these models
match as closely as possible the observed contact maps. These ``model-driven''
approaches offer mechanistical insights into the genome architecture.
